\documentclass[11pt]{article}
\usepackage{url}
\usepackage{alltt}
\usepackage{bm}
\linespread{1}
\textwidth 6.5in
\oddsidemargin 0.in
\addtolength{\topmargin}{-1in}
\addtolength{\textheight}{2in}

\usepackage{amsmath}
\usepackage{amssymb}

\begin{document}


\begin{center}
\Large
STA 712 Challenge Assignment 7: Poisson Regression in the Wild\\
\normalsize
\vspace{5mm}
\end{center}

\noindent \textbf{Due:} Monday, November 28, 12:00pm (noon) on Canvas\\ 

\noindent \textbf{Instructions:} 
\begin{itemize}
\item Submit your work as a single PDF. For this assignment, you may include written work by scanning it and incorporating it into the PDF. Include all R code needed to reproduce your results in your submission.
\item You are welcome to work with others on this assignment, but you must submit your own work.
\end{itemize}

\section*{Introduction}

Poisson regression (and variants like quasi-Poisson and ZIP models) is a popular choice for modeling count data, and is used in many applications. The purpose of this assignment is to see how Poisson regression is used by practitioners in real research projects. In this assignment, you will find a research paper which applies Poisson regression to real data. You will then write a short review of the paper, describing what the authors did, what their results were, and whether there were any issues in their methods. Finally, you will attempt to reproduce their results, using their data.

\section*{Questions}

\begin{enumerate}
\item Find a research paper in the sciences (biology, physics, chemistry, etc.) or the social sciences (psychology, sociology, etc.) which uses Poisson regression (or quasi-Poisson, or ZIP models) as a \textit{main} analysis method to address a research question. Make sure to choose a paper for which the data is publicly available; part of this assignment will be to reproduce the original analysis.

\item Write a short peer review of your chosen paper. The purpose of this review is to describe how the authors used Poisson regression in their analysis. Your review should answer the following questions:
\begin{enumerate}
\item What research question are the authors trying to answer?
\item What data do the authors collect to answer their research question?
\item What statistical methods do the authors use to analyze their data; in particular, how do the authors use Poisson regression to help answer the question?
\item Does their use of Poisson regression appear reasonable, or are their any issues with their analysis?
\item Is their anything you would suggest to improve the paper?
\end{enumerate}

\item Using the data associated with the paper, fit the same regression model as the authors. (If the paper fits several regression models, you may choose one). If the paper does not provide sufficient detail to reproduce their model, explain what additional detail you would need, then fit your best guess at their Poisson regression model.

\item Assess the Poisson regression assumptions using model diagnostics from class (empirical log mean plots, quantile residual plots, leverage and Cook's distance, and variance inflation factors).

\begin{enumerate}
\item Are there any issues with the fitted model? If so, do your best to address these issues (e.g. with variable transformations), and report your final model.

\item What diagnostics (if any) did the researchers use to assess the Poisson regression assumptions in their paper?
\end{enumerate}

\item Does your regression model support the researchers' conclusions in the original paper? Carry out any additional analysis necessary to answer the research question (e.g., hypothesis testing), and summarize your findings.

\end{enumerate}


\end{document}