---
title: Introduction to multinomial regression
output:
  xaringan::moon_reader:
    css: "lab-slides.css"
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
editor_options: 
  chunk_output_type: console
---

### Motivating example: earthquake data

.large[
We have data from the 2015 Gorkha earthquake in Nepal. After the earthquake, a large scale survey was conducted to determine the amount of damage the earthquake caused for homes, businesses and other structures. Variables include:
* `Damage`: the amount of damage suffered by the building (none, moderate, severe)
* `age`: the age of the building (in years)
* `condition`: a de-identified variable recording the condition of the land surrounding the building

**Research goal:** Build a model to predict Damage

.question[
Are any of the models we have learned so far suitable for predicting Damage?
]
]

---

### The categorical distribution

---

### Multivariate GLM

---

### Multinomial regression model

---

### Exploratory data analysis

.large[
.question[
We want to model damage using age and land surface condition. What kind of EDA could I do?
]
]

---

### Empirical logit plots

```{r, echo=F, message=F, warning=F, fig.width=10, fig.height=4}
library(tidyverse)
library(nnet)
library(gridExtra)
library(knitr)

hook_output <- knit_hooks$get("output")
knit_hooks$set(output = function(x, options) {
   lines <- options$output.lines
   if (is.null(lines)) {
     return(hook_output(x, options))  # pass to default hook
   }
   x <- unlist(strsplit(x, "\n"))
   more <- "..."
   if (length(lines)==1) {        # first n lines
     if (length(x) > lines) {
       # truncate the output, but add ....
       x <- c(head(x, lines), more)
     }
   } else {
     x <- c(more, x[lines], more)
   }
   # paste these lines together
   x <- paste(c(x, ""), collapse = "\n")
   hook_output(x, options)
 })

log_rr_plot <- function(data, num_bins, bin_method,
                        x, y, cat_num, cat_denom,
                        grouping = NULL, reg_formula = y ~ x){
  
  if(is.null(grouping)){
    dat <- data.frame(x = data[,x], 
                      y = data[,y],
                      group = 1)
  } else {
    dat <- data.frame(x = data[,x], 
                      y = data[,y],
                      group = data[,grouping])
  }
  
  dat <- dat %>%
    filter(y %in% c(cat_num, cat_denom)) %>%
    mutate(obs = ifelse(y == cat_num, 1, 0))
  
  
  if(bin_method == "equal_size"){
    logodds_table <- dat %>%
      slice_sample(n = nrow(dat), replace=F) %>%
      drop_na() %>%
      arrange(group, x) %>%
      group_by(group) %>%
      mutate(bin = rep(1:num_bins,
                       each=ceiling(n()/num_bins))[1:n()]) %>%
      group_by(bin, group) %>%
      summarize(mean_x = mean(x),
                prop = mean(c(obs, 0.5)),
                num_obs = n()) %>%
      ungroup() %>%
      mutate(logodds = log(prop/(1 - prop)))
  } else {
    logodds_table <- dat %>%
      slice_sample(n = nrow(dat), replace=F) %>%
      drop_na() %>%
      group_by(group) %>%
      mutate(bin = cut(x, 
                       breaks = num_bins,
                       labels = F)) %>%
      group_by(bin, group) %>%
      summarize(mean_x = mean(x),
                prop = mean(c(obs, 0.5)),
                num_obs = n()) %>%
      ungroup() %>%
      mutate(logodds = log(prop/(1 - prop)))
  }
  
  if(is.null(grouping)){
    logodds_table %>%
      ggplot(aes(x = mean_x,
                 y = logodds)) +
      geom_point(size=2.5) +
      geom_smooth(se=F, method="lm", formula = reg_formula) +
      theme_bw() +
      labs(x = x,
           y = paste("Log odds", cat_num, "vs.", cat_denom)) +
      theme(text = element_text(size=15))
  } else {
    logodds_table %>%
      ggplot(aes(x = mean_x,
                 y = logodds,
                 color = group,
                 shape = group)) +
      geom_point(size=2.5) +
      geom_smooth(se=F, method="lm", formula = reg_formula) +
      theme_bw() +
      labs(x = x,
           y = paste("Log odds", cat_num, "vs.", cat_denom),
           color = grouping,
           shape = grouping) +
      theme(text = element_text(size=15))
  }
  
}

earthquake <- read.csv("https://sta214-f22.github.io/labs/EarthquakeData.csv")

set.seed(100)

earthquake <- earthquake %>%
  slice_sample(n=10000) %>%
  dplyr::select(Damage, count_families, 
         foundation_type, land_surface_condition, 
         age) %>%
  mutate(Damage = factor(Damage, levels = c("none", "moderate", "severe"))) %>%
  rename(condition = land_surface_condition)

p1 <- earthquake %>%
  log_rr_plot(35, "equal_size", 
            x = "age", y = "Damage", 
            cat_num = "moderate", cat_denom = "none", 
            reg_formula = y ~ x) +
  theme(text = element_text(size=20))

p2 <- earthquake %>%
  log_rr_plot(35, "equal_size", 
            x = "age", y = "Damage", 
            cat_num = "severe", cat_denom = "none", 
            reg_formula = y ~ x) +
  theme(text = element_text(size=20))

grid.arrange(p1, p2, ncol=2)
```

---

### Trying a transformation

```{r, echo=F, message=F, warning=F, fig.align='center', fig.width=10, fig.height=7.5}
p1 <- earthquake %>%
  log_rr_plot(35, "equal_size", 
            x = "age", y = "Damage", 
            cat_num = "moderate", cat_denom = "none", 
            reg_formula = y ~ x) 

p2 <- earthquake %>%
  log_rr_plot(35, "equal_size", 
            x = "age", y = "Damage", 
            cat_num = "severe", cat_denom = "none", 
            reg_formula = y ~ x)

p3 <- earthquake %>%
  mutate(sqrt_age = sqrt(age)) %>%
  log_rr_plot(35, "equal_size", 
            x = "sqrt_age", y = "Damage", 
            cat_num = "moderate", cat_denom = "none", 
            reg_formula = y ~ x) +
  labs(x = "sqrt(age)")

p4 <- earthquake %>%
  mutate(sqrt_age = sqrt(age)) %>%
  log_rr_plot(35, "equal_size", 
            x = "sqrt_age", y = "Damage", 
            cat_num = "severe", cat_denom = "none", 
            reg_formula = y ~ x) +
  labs(x = "sqrt(age)")

grid.arrange(p1, p2, p3, p4, ncol=2)
```

---

### Fitting the model in R

```{r, message=F, warning=F, results='hide'}
library(nnet)
m1 <- multinom(Damage ~ sqrt(age) + 
                 condition, 
               data = earthquake)
```

```{r, output.lines = 4:12}
summary(m1)
```

---

### Class activity

.large[
[https://sta712-f22.github.io/class_activities/ca_lecture_34.html](https://sta712-f22.github.io/class_activities/ca_lecture_34.html)
]