---
title: Multinomial regression
output:
  xaringan::moon_reader:
    css: "lab-slides.css"
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
editor_options: 
  chunk_output_type: console
---

### Motivating example: earthquake data

.large[
We have data from the 2015 Gorkha earthquake in Nepal. After the earthquake, a large scale survey was conducted to determine the amount of damage the earthquake caused for homes, businesses and other structures. Variables include:
* `Damage`: the amount of damage suffered by the building (none, moderate, severe)
* `age`: the age of the building (in years)
* `condition`: a de-identified variable recording the condition of the land surrounding the building

]

---

### Fisher scoring

---

### Wald tests

```{r, include=F}
library(tidyverse)
library(nnet)
library(gridExtra)
library(knitr)

hook_output <- knit_hooks$get("output")
knit_hooks$set(output = function(x, options) {
   lines <- options$output.lines
   if (is.null(lines)) {
     return(hook_output(x, options))  # pass to default hook
   }
   x <- unlist(strsplit(x, "\n"))
   more <- "..."
   if (length(lines)==1) {        # first n lines
     if (length(x) > lines) {
       # truncate the output, but add ....
       x <- c(head(x, lines), more)
     }
   } else {
     x <- c(more, x[lines], more)
   }
   # paste these lines together
   x <- paste(c(x, ""), collapse = "\n")
   hook_output(x, options)
 })

earthquake <- read.csv("~/Documents/Teaching/sta712-f22.github.io/slides/earthquake_small.csv") %>%
  mutate(Damage = factor(Damage, levels = c("none", "moderate", "severe")))

m1 <- multinom(Damage ~ sqrt(age) + condition,
               data = earthquake)

```

.large[
```{r, echo=F, message=F, warning=F, output.lines=4:12}
summary(m1)
```
]

.large[
.question[
Suppose we want to know whether there is a relationship between age and the odds of moderate vs. no damage, after accounting for surface condition. What hypotheses would we test?
]
]

---

### Wald tests

.large[
```{r, echo=F, message=F, warning=F, output.lines=4:12}
summary(m1)
```
]

.large[
.question[
Suppose we want to know whether the relationship between age and the odds of moderate vs. no damage is the *same* as the relationship between age and the odds of severe vs. no damage. What hypotheses would we test?
]
]

---

### Wald tests

.large[
```{r}
diff <- t(c(0, -1, 0, 0, 0, 1, 0, 0)) %*% 
  c(t(coef(m1)))
std_err <- sqrt(t(c(0, -1, 0, 0, 0, 1, 0, 0)) %*% 
                  vcov(m1) %*% 
                  c(0, -1, 0, 0, 0, 1, 0, 0))
(diff - 0)/std_err

2*pnorm((diff - 0)/std_err, lower.tail = F)
```
]

---

### Likelihood ratio tests

.large[
```{r, echo=F, message=F, warning=F, output.lines=4:12}
summary(m1)
```
]

.large[
.question[
Suppose we want to know whether there is a relationship between surface condition and damage, after accounting for building age. What hypotheses would we test?
]
]

---

### Likelihood ratio tests

.large[
```{r, results = F}
m1 <- multinom(Damage ~ sqrt(age) + condition,
               data = earthquake)
m2 <- multinom(Damage ~ sqrt(age),
               data = earthquake)

pchisq(m2$deviance - m1$deviance, df = 4, 
       lower.tail=F)
```

```{r, echo=F}
pchisq(m2$deviance - m1$deviance, df = 4, lower.tail=F)
```
]

---

### Class activity

.large[
[https://sta712-f22.github.io/class_activities/ca_lecture_36.html](https://sta712-f22.github.io/class_activities/ca_lecture_36.html)
]