---
title: Likelihood ratio tests
output:
  xaringan::moon_reader:
    css: "lab-slides.css"
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
editor_options: 
  chunk_output_type: console
---

### Last time

.large[
Data on the RMS *Titanic* disaster. We have data on 891 passengers on the ship, with the following variables:

* `Passenger`: A unique ID number for each passenger.
* `Survived`: An indicator for whether the passenger survived (1) or perished (0) during the disaster.
* `Pclass`: Indicator for the class of the ticket held by this passengers; 1 = 1st class, 2 = 2nd class, 3 = 3rd class.
* `Sex`: Binary Indicator for the biological sex of the passenger.
* `Age`: Age of the passenger in years; Age is fractional if the passenger was less than 1 year old.
* `Fare`: How much the ticket cost in US dollars.
* `+` others
]

---

### Last time

.large[
*Is there a relationship between passenger age and their probability of survival, after accounting for sex, passenger class, and the cost of their ticket?*
]

.large[
$$\log \left( \dfrac{p_i}{1 - p_i} \right) = \beta_0 + \beta_1 Age_i + \beta_2 Sex_i + \\ \hspace{2cm} \beta_3 Age_i \cdot Sex_i + \beta_4 \log(Fare_i + 1)$$
]

.large[
.question[
What hypotheses should we test to investigate this research question?
]
]

---

### Likelihood ratio tests

```{r include=F}
library(knitr)
library(tidyverse)
hook_output <- knit_hooks$get("output")
knit_hooks$set(output = function(x, options) {
   lines <- options$output.lines
   if (is.null(lines)) {
     return(hook_output(x, options))  # pass to default hook
   }
   x <- unlist(strsplit(x, "\n"))
   more <- "..."
   if (length(lines)==1) {        # first n lines
     if (length(x) > lines) {
       # truncate the output, but add ....
       x <- c(head(x, lines), more)
     }
   } else {
     x <- c(more, x[lines], more)
   }
   # paste these lines together
   x <- paste(c(x, ""), collapse = "\n")
   hook_output(x, options)
 })

titanic <- read.csv("https://sta214-f22.github.io/labs/Titanic.csv") %>%
  drop_na()
```

.large[
```{r, echo=F, output.lines = c(10:16, 21:23)}
m1 <- glm(Survived ~ Age*Sex + log(Fare + 1), data = titanic, 
          family = binomial)
summary(m1)
```
]

.large[
.question[
What information replaces $R^2$ and $R^2_{adj}$ in the GLM output?
]
]

---

### Deviance

.large[
**Definition:** The *deviance* of a fitted model with parameter estimates $\widehat{\beta}$ is given by
$$2 \ell(\text{saturated model}) - 2 \ell(\widehat{\beta})$$
]

---

### Residual and null deviance

.large[
```{r, output.lines = c(22:23)}
m1 <- glm(Survived ~ Age*Sex + log(Fare + 1), 
          data = titanic, family = binomial)
summary(m1)
```
]

---

### Comparing deviances

.large[
```{r, output.lines = c(22:23)}
m1 <- glm(Survived ~ Age*Sex + log(Fare + 1), 
          data = titanic, family = binomial)
summary(m1)
```
]

.large[
```{r, output.lines = c(20:21)}
m2 <- glm(Survived ~ Sex + log(Fare + 1), 
          data = titanic, family = binomial)
summary(m2)
```
]

---

### Comparing deviances

.large[
**Full model:** 

</br>

**Hypotheses:** 

</br>

**Reduced model:**

</br>

**Test statistic:**

]

---

### Comparing deviances

.large[
**Full model:** $\log \left( \dfrac{p_i}{1 - p_i} \right) = \beta_0 + \beta_1 Sex_i + \beta_2 \log(Fare_i + 1) + \\ \hspace{3cm} \beta_3 Age_i + \beta_4 Age_i \cdot Sex_i$

**Reduced model:** $\log \left( \dfrac{p_i}{1 - p_i} \right) = \beta_0 + \beta_1 Sex_i + \beta_2 \log(Fare_i + 1)$

$$G = 2  \ell(\widehat{\beta}) - 2 \ell(\widehat{\beta}^0)$$

.question[
Why is $G$ always $\geq 0$?
]
]

---

### Comparing deviances

.large[
**Full model:** $\log \left( \dfrac{p_i}{1 - p_i} \right) = \beta_0 + \beta_1 Sex_i + \beta_2 \log(Fare_i + 1) + \\ \hspace{3cm} \beta_3 Age_i + \beta_4 Age_i \cdot Sex_i$

**Reduced model:** $\log \left( \dfrac{p_i}{1 - p_i} \right) = \beta_0 + \beta_1 Sex_i + \beta_2 \log(Fare_i + 1)$

$$G = 2 \ell(\widehat{\beta}) - 2 \ell(\widehat{\beta}^0) = 10.83$$

.question[
If the reduced model is correct, how unusual is $G = 10.83$?
]
]

---

### Likelihood ratio test

