---
title: Confidence intervals
output:
  xaringan::moon_reader:
    css: "lab-slides.css"
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
editor_options: 
  chunk_output_type: console
---

```{r include=F}
library(knitr)
library(tidyverse)
hook_output <- knit_hooks$get("output")
knit_hooks$set(output = function(x, options) {
   lines <- options$output.lines
   if (is.null(lines)) {
     return(hook_output(x, options))  # pass to default hook
   }
   x <- unlist(strsplit(x, "\n"))
   more <- "..."
   if (length(lines)==1) {        # first n lines
     if (length(x) > lines) {
       # truncate the output, but add ....
       x <- c(head(x, lines), more)
     }
   } else {
     x <- c(more, x[lines], more)
   }
   # paste these lines together
   x <- paste(c(x, ""), collapse = "\n")
   hook_output(x, options)
 })

titanic <- read.csv("https://sta214-f22.github.io/labs/Titanic.csv")

titanic <- titanic %>%
  drop_na() %>%
  mutate(Pclass = factor(Pclass, levels = c(3, 2, 1))) 
```

### Last time: Wald confidence intervals

---

### Example: Titanic data

.large[
$$\log \left( \dfrac{p_i}{1 - p_i} \right) = \beta_0 + \beta_1 Sex_i + \beta_2 Age_i + \beta_3 SecondClass_i + \\ \hspace{3cm} \beta_4 FirstClass_i + \beta_5 Sex_i \cdot Age_i$$

```{r echo=F, output.lines = 10:17}
m1 <- glm(Survived ~ Sex*Age + Pclass, data = titanic, family = binomial)
summary(m1)
```

]

---

### Confidence intervals for linear combinations

---

### Class activity

.large[
[https://sta712-f22.github.io/class_activities/ca_lecture_16.html](https://sta712-f22.github.io/class_activities/ca_lecture_16.html)
]

---

### Inverting the likelihood ratio test

---

### Types of research questions

.large[
So far, we have learned how to answer the following questions:

* What is the relationship between the explanatory variable(s) and the response?
* What is a "reasonable range" for a parameter in this relationship?
* Do we have strong evidence for a relationship between these variables?

.question[
What other kinds of research questions might we ask?
]
]

---

### Making predictions

.large[
* For each passenger, we calculate $\widehat{p}_i$ (estimated probability of survival)
* But, we want to predict *which* passengers actually survive

.question[
How do we turn $\widehat{p}_i$ into a binary prediction of survival / no survival?
]
]

---

### Confusion matrix

.large[
| | | Actual | |
| --- | --- | --- | --- |
| | | $Y = 0$ | $Y = 1$ |
|**Predicted** | $\widehat{Y} = 0$ | 344 | 70 |
| | $\widehat{Y} = 1$ | 80 | 220 |
]

.large[
.question[
Did we do a good job predicting survival?
]
]