---
title: Overdispersion
output:
  xaringan::moon_reader:
    css: "lab-slides.css"
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
editor_options: 
  chunk_output_type: console
---

```{r include=F}
library(tidyverse)
library(knitr)
hook_output <- knit_hooks$get("output")
knit_hooks$set(output = function(x, options) {
   lines <- options$output.lines
   if (is.null(lines)) {
     return(hook_output(x, options))  # pass to default hook
   }
   x <- unlist(strsplit(x, "\n"))
   more <- "..."
   if (length(lines)==1) {        # first n lines
     if (length(x) > lines) {
       # truncate the output, but add ....
       x <- c(head(x, lines), more)
     }
   } else {
     x <- c(more, x[lines], more)
   }
   # paste these lines together
   x <- paste(c(x, ""), collapse = "\n")
   hook_output(x, options)
 })

crimes <- read_csv("~/Documents/Teaching/sta712-f22.github.io/slides/c_data.csv")
```

### Data

.large[
A concerned parent asks us to investigate crime rates on college campuses. We have access to data on 81 different colleges and universities in the US, including the following variables:

* `type`: college (C) or university (U)
* `nv`: the number of violent crimes for that institution in the given year
* `enroll1000`: the number of enrolled students, in thousands
* `region`: region of the US C = Central, MW = Midwest, NE = Northeast, SE = Southeast, SW = Southwest, and W = West)
]

---

### Offsets

.large[
We will account for school size by including an **offset** in the model:

$$Crimes_i \sim Poisson(\lambda_i)$$

$$\log(\lambda_i) = \beta_0 + \beta_1 MW_i + \beta_2 NE_i + \beta_3 SE_i + \beta_4 SW_i + \beta_5 W_i \\ + \log(Enrollment_i)$$
]

---

### Fitting a model with an offset

.large[
```{r, output.lines = 10:16}
m2 <- glm(nv ~ region, offset = log(enroll1000),
          data = crimes, family = poisson)
summary(m2)
```

* The offset doesn't show up in the output (because we're not estimating a coefficient for it)
]

---

### Fitting a model with an offset

.large[
$$\log(\widehat{\lambda}_i) = -1.30 + 0.10 MW_i + 0.76 NE_i + \\ 0.87 SE_i + 0.51 SW_i + 0.21 W_i \\  + \log(Enrollment_i)$$

.question[
How would I interpret the intercept -1.30?
]
]

---

### When to use offsets

.large[
Offsets are useful in Poisson regression when our counts come from groups of very different sizes (e.g., different numbers of students on a college campus). The offset lets us interpret model coefficients in terms of rates instead of raw counts.

.question[
With your neighbor, brainstorm some other data scenarios where our response is a count variable, and an offset would be useful. What would our offset be?
]
]

---

### Goodness of fit

.large[
```{r, output.lines = 20:23}
m2 <- glm(nv ~ region, offset = log(enroll1000),
          data = crimes, family = poisson)
summary(m2)
```

```{r}
pchisq(433.14, df=75, lower.tail=F)
```
]

---

### Overdispersion

.large[
**Overdispersion** occurs when the response $Y$ has higher variance than we would expect from the specified EDM

.question[
Why is it a problem if $Y$ has more variance than we account for in our model?
]
]

---

### Estimating $\phi$

---

### Using $\widehat{\phi}$

.large[
```{r}
pearson_resids <- residuals(m2, type="pearson")
sum(pearson_resids^2)/df.residual(m2)
```

```{r echo=F, output.lines = 10:16}
summary(m2)
```
]
