---
title: Confidence intervals
output:
  xaringan::moon_reader:
    css: "lab-slides.css"
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
editor_options: 
  chunk_output_type: console
---

```{r include=F}
library(knitr)
library(tidyverse)
hook_output <- knit_hooks$get("output")
knit_hooks$set(output = function(x, options) {
   lines <- options$output.lines
   if (is.null(lines)) {
     return(hook_output(x, options))  # pass to default hook
   }
   x <- unlist(strsplit(x, "\n"))
   more <- "..."
   if (length(lines)==1) {        # first n lines
     if (length(x) > lines) {
       # truncate the output, but add ....
       x <- c(head(x, lines), more)
     }
   } else {
     x <- c(more, x[lines], more)
   }
   # paste these lines together
   x <- paste(c(x, ""), collapse = "\n")
   hook_output(x, options)
 })

titanic <- read.csv("https://sta214-f22.github.io/labs/Titanic.csv")

titanic <- titanic %>%
  drop_na() %>%
  mutate(Pclass = factor(Pclass, levels = c(3, 2, 1))) 
```

### Announcements

.large[
* HW 4 released, due next Friday
* Exam 1 released next Friday
  * Take home, 1 week to complete
  * Open note (anything from this course)
  * Closed internet
  * Closed other people
* Reminder: department seminar on Monday, 12pm - 1pm (Dr. Mine Cetinkaya-Rundel)
  * Can sign up to meet with the speaker 11 - 11:30
]

---

### Wald vs. likelihood ratio tests

---

### Confidence intervals

.large[
$$\log \left( \dfrac{p_i}{1 - p_i} \right) = \beta_0 + \beta_1 Sex_i + \beta_2 Age_i + \beta_3 SecondClass_i + \\ \hspace{3cm} \beta_4 FirstClass_i + \beta_5 Sex_i \cdot Age_i$$

```{r echo=F, output.lines = 10:17}
m1 <- glm(Survived ~ Sex*Age + Pclass, data = titanic, family = binomial)
summary(m1)
```

.question[
How do I create a 95% confidence interval for $\beta_3$?
]
]

---

### Wald confidence intervals

---

### Confidence intervals for linear combinations

---

### Class activity

.large[
[https://sta712-f22.github.io/class_activities/ca_lecture_15.html](https://sta712-f22.github.io/class_activities/ca_lecture_15.html)
]

