---
title: Logistic regression assumptions and diagnostics
output:
  xaringan::moon_reader:
    css: "lab-slides.css"
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
editor_options: 
  chunk_output_type: console
---

### Motivating example: Dengue data

.large[
**Data:** Data on 5720 Vietnamese children, admitted to the hospital with possible dengue fever. Variables include:

* *Sex*: patient's sex (female or male)
* *Age*: patient's age (in years)
* *WBC*: white blood cell count
* *PLT*: platelet count
* other diagnostic variables...
* *Dengue*: whether the patient has dengue (0 = no, 1 = yes)
]

---

### Previously: Logistic regression model

.large[
$$Y_i = \text{dengue status (0 = negative, 1 = positive)}$$

$$Y_i \sim Bernoulli(p_i)$$

$$\log \left( \dfrac{p_i}{1 - p_i} \right) = \beta_0 + \beta_1 WBC_i$$

.question[
What assumptions does this logistic regression model make? How should we assess these assumptions? Discuss with your neighbor for 2--3 minutes, then we will discuss as a group.
]
]

---

### Don't use usual residuals for logistic regression

.large[
Fitted model: $\log \left(\dfrac{\widehat{p}_i}{1 - \widehat{p}_i}\right) = 1.737 - 0.361 \ WBC_i$
]

.large[
Residuals $Y_i - \widehat{p}_i$:
]

```{r echo=F, message=F, fig.width=7, fig.height = 5, fig.align='center'}
library(tidyverse)
dengue <- read.csv("https://sta279-s22.github.io/labs/dengue.csv")

m1 <- glm(Dengue ~ WBC, data = dengue, family = binomial)
data.frame(WBC = dengue$WBC,
           Residual = dengue$Dengue - m1$fitted.values) %>%
  ggplot(aes(x = WBC, y = Residual)) +
  geom_point(size=1.5) +
  geom_abline(slope = 0, intercept = 0, color="blue", lwd=1.2) +
  theme_bw() +
  theme(text = element_text(size = 20))
```

---

### Assessing shape with empirical logit plots

.large[
**Example:** Putting data. Interested in the relationship between the length of a putt, and whether it was made:

$$Y_i \sim Bernoulli(p_i)$$

$$\log \left( \dfrac{p_i}{1 - p_i} \right) = \beta_0 + \beta_1 \ Length_i$$

| Length | 3 | 4 | 5 | 6 | 7 |
| --- | --- | --- | --- | --- | --- |
| Number of successes | 84 | 88 | 61 | 61 | 44 |
| Number of failures | 17 | 31 | 47 | 64 | 90 |
| Total | 101 | 119 | 108 | 125 | 134 |
]

---

### Empirical logits

.large[
**Step 1:** estimate the probability of success for each length of putt

| Length | 3 | 4 | 5 | 6 | 7 |
| --- | --- | --- | --- | --- | --- |
| Number of successes | 84 | 88 | 61 | 61 | 44 |
| Number of failures | 17 | 31 | 47 | 64 | 90 |
| Total | 101 | 119 | 108 | 125 | 134 |
| Probability of success $\widehat{p}$ | 0.832 | 0.739 | 0.565 | 0.488 | 0.328 |
]

---

### Empirical logits

.large[
**Step 2:** convert empirical probabilities to empirical log odds

| Length | 3 | 4 | 5 | 6 | 7 |
| --- | --- | --- | --- | --- | --- |
| Number of successes | 84 | 88 | 61 | 61 | 44 |
| Number of failures | 17 | 31 | 47 | 64 | 90 |
| Total | 101 | 119 | 108 | 125 | 134 |
| Probability of success $\widehat{p}$ | 0.832 | 0.739 | 0.565 | 0.488 | 0.328 |
| Odds $\dfrac{\widehat{p}}{1 - \widehat{p}}$ | 4.941 | 2.839 | 1.298 | 0.953 | 0.489 |
| Log-odds $\log \left( \dfrac{\widehat{p}}{1 - \widehat{p}} \right)$ | 1.60 | 1.04 | 0.26 | -0.05 | -0.72 |
]

---

### Empirical logits

.large[
**Step 3:** plot empirical log-odds against predictor, and add a least-squares line

```{r echo=F, message=F, fig.width=7, fig.height=4.5, fig.align='center'}
data.frame(length = c(3, 4, 5, 6, 7),
  log_odds = c(1.6, 1.04, 0.26, -0.05, -0.72)) %>% 
  ggplot(aes(x = length, y = log_odds)) + 
  geom_point(size=3) + 
  geom_smooth(se=F, method="lm") +
  labs(x = "Length (feet)",
       y = "Empirical log-odds") +
  theme_bw() +
  theme(text = element_text(size = 25))
```

.question[
Does it seem reasonable that the log-odds are a linear function of length?
]
]

---

### Back to the dengue data...

.large[

| WBC | 0.90 | 1.15 | 1.23 | 1.25 | 1.54 | 1.58 | ... |
| --- | --- | --- | --- | --- | --- | --- | --- |
| Dengue = 0 | 0 | 0 | 0 | 0 | 0 | 0 | ... |
| Dengue = 1 | 1 | 2 | 1 | 1 | 3 | 1 | ... |

.question[
What problem do I run into?
]
]

---

### Binned empirical logit plots

```{r echo=F, message=F, fig.align='center', fig.width=7, fig.height=5}
logodds_plot <- function(data, num_bins, bin_method,
                         x_name, y_name, grouping = NULL, reg_formula = y ~ x){
  
  if(is.null(grouping)){
    dat <- data.frame(x = data %>% pull(x_name), 
                      y = data %>% pull(y_name),
                      group = 1)
  } else {
    dat <- data.frame(x = data %>% pull(x_name), 
                      y = data %>% pull(y_name),
                      group = data %>% pull(grouping))
  }
  
  if(bin_method == "equal_size"){
    logodds_table <- dat %>%
      drop_na() %>%
      arrange(group, x) %>%
      group_by(group) %>%
      mutate(obs = y,
             bin = rep(1:num_bins,
                       each=ceiling(n()/num_bins))[1:n()]) %>%
      group_by(bin, group) %>%
      summarize(mean_x = mean(x),
                prop = mean(c(obs, 0.5)),
                num_obs = n()) %>%
      ungroup() %>%
      mutate(logodds = log(prop/(1 - prop)))
  } else {
    logodds_table <- dat %>%
      drop_na() %>%
      group_by(group) %>%
      mutate(obs = y,
             bin = cut(x, 
                       breaks = num_bins,
                       labels = F)) %>%
      group_by(bin, group) %>%
      summarize(mean_x = mean(x),
                prop = mean(c(obs, 0.5)),
                num_obs = n()) %>%
      ungroup() %>%
      mutate(logodds = log(prop/(1 - prop)))
  }
  
  if(is.null(grouping)){
    logodds_table %>%
      ggplot(aes(x = mean_x,
                 y = logodds)) +
      geom_point(size=2) +
      geom_smooth(se=F, method="lm", formula = reg_formula) +
      theme_bw() +
      labs(x = x_name,
           y = "Empirical log odds") +
      theme(text = element_text(size=25))
  } else {
    logodds_table %>%
      ggplot(aes(x = mean_x,
                 y = logodds,
                 color = group,
                 shape = group)) +
      geom_point(size=2) +
      geom_smooth(se=F, method="lm", formula = reg_formula) +
      theme_bw() +
      labs(x = x_name,
           y = "Empirical log odds",
           color = grouping,
           shape = grouping) +
      theme(text = element_text(size=25))
  }
  
}

dengue %>%
  logodds_plot(num_bins = 15, bin_method = "equal_size", 
               x_name = "WBC", y_name = "Dengue",
               reg_formula = y ~ x)
```

---

### Binned empirical logit plots

```{r echo=F, message=F, fig.align='center', fig.width=7, fig.height=5}
dengue %>%
  logodds_plot(num_bins = 15, bin_method = "equal_size", 
               x_name = "WBC", y_name = "Dengue",
               reg_formula = y ~ x)
```

.large[
.question[
Does it seem reasonable that the log-odds are a linear function of WBC?
]
]

---

### Trying some transformations

```{r echo=F, message=F, fig.align='center', fig.width=10, fig.height=7}
library(gridExtra)
p1 <- dengue %>% 
  logodds_plot(num_bins = 15, bin_method = "equal_size", 
               x_name = "WBC", y_name = "Dengue",
               reg_formula = y ~ x) +
  labs(title = "Untransformed")

p2 <- dengue %>% 
  logodds_plot(num_bins = 15, bin_method = "equal_size", 
               x_name = "WBC", y_name = "Dengue",
               reg_formula = y ~ log(x)) +
  labs(title = "Log transformation")

p3 <- dengue %>% 
  logodds_plot(num_bins = 15, bin_method = "equal_size", 
               x_name = "WBC", y_name = "Dengue",
               reg_formula = y ~ sqrt(x)) +
  labs(title = "Square root")

p4 <- dengue %>% 
  logodds_plot(num_bins = 15, bin_method = "equal_size", 
               x_name = "WBC", y_name = "Dengue",
               reg_formula = y ~ I(1/x)) +
  labs(title = "Reciprocal ")

grid.arrange(p1, p2, p3, p4, ncol=2)
```

---

### Why residuals in linear regression are nice

---

### Quantile residuals for logistic regression

---

### Quantile residuals example (in R)

---

### Class activity, Part I

.large[
[https://sta712-f22.github.io/class_activities/ca_lecture_5.html](https://sta712-f22.github.io/class_activities/ca_lecture_5.html)
]

---

### Class activity, Part II

.large[
[https://sta712-f22.github.io/class_activities/ca_lecture_5.html](https://sta712-f22.github.io/class_activities/ca_lecture_5.html)
]