---
title: Wald tests and likelihood ratio tests
output:
  xaringan::moon_reader:
    css: "lab-slides.css"
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
editor_options: 
  chunk_output_type: console
---


### Wald tests for multiple parameters

---

### Class activity

.large[
[https://sta712-f22.github.io/class_activities/ca_lecture_11.html](https://sta712-f22.github.io/class_activities/ca_lecture_11.html)

.question[
* Wald tests for the dengue data
]
]

---

### Likelihood ratio tests

```{r include=F}
library(knitr)
hook_output <- knit_hooks$get("output")
knit_hooks$set(output = function(x, options) {
   lines <- options$output.lines
   if (is.null(lines)) {
     return(hook_output(x, options))  # pass to default hook
   }
   x <- unlist(strsplit(x, "\n"))
   more <- "..."
   if (length(lines)==1) {        # first n lines
     if (length(x) > lines) {
       # truncate the output, but add ....
       x <- c(head(x, lines), more)
     }
   } else {
     x <- c(more, x[lines], more)
   }
   # paste these lines together
   x <- paste(c(x, ""), collapse = "\n")
   hook_output(x, options)
 })

dengue <- read.csv("https://sta712-f22.github.io/homework/dengue.csv")
```

.large[
```{r, output.lines = c(10:14, 19:21)}
m1 <- glm(Dengue ~ WBC + PLT, data = dengue, 
          family = binomial)
summary(m1)
```
]

.large[
.question[
What information replaces $R^2$ and $R^2_{adj}$ in the GLM output?
]
]

---

### Deviance

.large[
**Definition:** The *deviance* of a fitted model with parameter estimates $\widehat{\beta}$ is given by
$$2 \log \ell(\text{saturated model}) - 2 \log \ell(\widehat{\beta})$$
]

---

### Comparing deviances

.large[
```{r, output.lines = c(19:21)}
m1 <- glm(Dengue ~ WBC + PLT, data = dengue, 
          family = binomial)
summary(m1)
```
]

---

### Comparing deviances

.large[
**Full model:** $\log \left( \dfrac{p_i}{1 - p_i} \right) = \beta_0 + \beta_1 WBC_i + \beta_2 PLT_i$

**Reduced model:** $\log \left( \dfrac{p_i}{1 - p_i} \right) = \beta_0$

$$G = 2 \log \ell(\widehat{\beta}) - 2 \log \ell(\widehat{\beta}^0)$$

.question[
Why is $G$ always $\geq 0$?
]
]

---

### Comparing deviances

.large[
**Full model:** $\log \left( \dfrac{p_i}{1 - p_i} \right) = \beta_0 + \beta_1 WBC_i + \beta_2 PLT_i$

**Reduced model:** $\log \left( \dfrac{p_i}{1 - p_i} \right) = \beta_0$

$$G = 2 \log \ell(\widehat{\beta}) - 2 \log \ell(\widehat{\beta}^0) = 1556.1$$

.question[
If the reduced model is correct, how unusual is $G = 1556.1$?
]
]

---

### Likelihood ratio test

