---
title: Wald tests
output:
  xaringan::moon_reader:
    css: "lab-slides.css"
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
editor_options: 
  chunk_output_type: console
---

### Proof: asymptotic distribution of the MLE

---

### Wald tests for single parameters

.large[
Logistic regression model for the dengue data:

$$Y_i \sim Bernoulli(p_i)$$
]

.large[
$$\log \left( \dfrac{p_i}{1 - p_i} \right) = \beta_0 + \beta_1 WBC_i + \beta_2 PLT_i$$

.question[
Researchers want to know if there is a relationship between white blood cell count and the probability a patient has dengue, after accounting for platelet count. What hypotheses should the researchers test?
]
]

---

### Wald tests for single parameters

```{r include=F}
library(knitr)
hook_output <- knit_hooks$get("output")
knit_hooks$set(output = function(x, options) {
   lines <- options$output.lines
   if (is.null(lines)) {
     return(hook_output(x, options))  # pass to default hook
   }
   x <- unlist(strsplit(x, "\n"))
   more <- "..."
   if (length(lines)==1) {        # first n lines
     if (length(x) > lines) {
       # truncate the output, but add ....
       x <- c(head(x, lines), more)
     }
   } else {
     x <- c(more, x[lines], more)
   }
   # paste these lines together
   x <- paste(c(x, ""), collapse = "\n")
   hook_output(x, options)
 })

dengue <- read.csv("https://sta712-f22.github.io/homework/dengue.csv")
```

.large[
```{r, output.lines = 10:14}
m1 <- glm(Dengue ~ WBC + PLT, data = dengue, 
          family = binomial)
summary(m1)
```
]

---

### Wald tests for multiple parameters

.large[
Logistic regression model for the dengue data:

$$Y_i \sim Bernoulli(p_i)$$
]

.large[
$$\log \left( \dfrac{p_i}{1 - p_i} \right) = \beta_0 + \beta_1 WBC_i + \beta_2 PLT_i$$

.question[
Researchers want to know if there is any relationship between white blood cell count or platelet count, and the probability a patient has dengue. What hypotheses should they test?
]
]

---

### Wald tests for multiple parameters

.large[
```{r, output.lines = 10:14}
m1 <- glm(Dengue ~ WBC + PLT, data = dengue, 
          family = binomial)
summary(m1)
```
]

.large[
.question[
Can the researchers test their hypotheses using this output?
]
]

---

### Wald tests for multiple parameters

---

### Class activity

.large[
[https://sta712-f22.github.io/class_activities/ca_lecture_10.html](https://sta712-f22.github.io/class_activities/ca_lecture_10.html)

.question[
* Wald tests for the dengue data
]
]

