---
title: Negative binomial regression
output:
  xaringan::moon_reader:
    css: "lab-slides.css"
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
editor_options: 
  chunk_output_type: console
---

```{r include=F}
library(tidyverse)
library(knitr)
library(gridExtra)
library(statmod)
hook_output <- knit_hooks$get("output")
knit_hooks$set(output = function(x, options) {
   lines <- options$output.lines
   if (is.null(lines)) {
     return(hook_output(x, options))  # pass to default hook
   }
   x <- unlist(strsplit(x, "\n"))
   more <- "..."
   if (length(lines)==1) {        # first n lines
     if (length(x) > lines) {
       # truncate the output, but add ....
       x <- c(head(x, lines), more)
     }
   } else {
     x <- c(more, x[lines], more)
   }
   # paste these lines together
   x <- paste(c(x, ""), collapse = "\n")
   hook_output(x, options)
 })

heart_data <- read.csv("~/Documents/Teaching/sta214-f22.github.io/class_activities/framingham.csv")

heart_data <- heart_data %>%
  drop_na(male, age, education, diabetes, cigsPerDay, BMI) %>%
  mutate(education = as.factor(education))

smokers <- heart_data %>%
  filter(currentSmoker == 1)

m1 <- glm(cigsPerDay ~ male + age + education + diabetes + BMI,
          data = smokers, family = poisson)

library(MASS)
m2 <- glm.nb(cigsPerDay ~ male + age + education + 
               diabetes + BMI, data = smokers)

```

### Recap: inference with negative binomial models

.large[
```{r echo=F, output.lines = c(11:19)}
summary(m2)
```
]

---

### Likelihood ratio test

.large[
```{r}
m2 <- glm.nb(cigsPerDay ~ male + age + education + 
               diabetes + BMI, data = smokers)
m3 <- glm.nb(cigsPerDay ~ male + age +
               diabetes + BMI, data = smokers)
m2$twologlik - m3$twologlik
pchisq(1.423, df=3, lower.tail=F)
```
]

---

### Likelihood ratio test

.large[
.question[
Why can I use the residual deviance to perform a likelihood ratio test for a  Poisson regression model, but not for a negative binomial model?
]
]

---

### New data

.large[
Survey data from 77 college students on a dry campus (i.e., alcohol is prohibited) in the US. Survey asks students "How many alcoholic drinks did you consume last weekend?"

* `drinks`: the number of drinks the student reports consuming
* `sex`: an indicator for whether the student identifies as male
* `OffCampus`: an indicator for whether the student lives off campus
* `FirstYear`: an indicator for whether the student is a first-year student

.question[
Our goal: model the number of drinks students report consuming.
]
]

---

### EDA: drinks

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=7, fig.height=5}
library(tidyverse)

wdrinks <- read.csv("~/Documents/Teaching/sta279-s22.github.io/slides/weekendDrinks.csv")

FirstYear <- ifelse( wdrinks$dorm %in% c("mohn","kittlesby", "kildahl"), "TRUE", "FALSE")
OffCampus <- ifelse( wdrinks$dorm == "off campus", "TRUE", "FALSE")

wdrinks <- cbind(wdrinks, FirstYear, OffCampus)

wdrinks %>%
  ggplot(aes(x = drinks)) +
  geom_bar() +
  theme_bw() +
  theme(text = element_text(size = 25))
```

.large[
.question[
What do you notice about this distribution?
]
]

---

### Comparisons with Poisson distributions

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=10, fig.height=7}
library(gridExtra)
set.seed(1)
p1 <- wdrinks %>%
  ggplot(aes(x = drinks)) +
  geom_bar() +
  theme_bw() +
  theme(text = element_text(size = 25)) +
  labs(title = "Observed data", x = "", y = "") +
  scale_x_continuous(limits = c(-1, 25)) +
  scale_y_continuous(limits = c(0, 45))

p2 <- data.frame(x = rpois(100, 1)) %>%
  ggplot(aes(x = x)) +
  geom_bar() +
  theme_bw() +
  theme(text = element_text(size = 25)) +
  labs(title = "Poisson(1)", x = "", y = "") +
  scale_x_continuous(limits = c(-1, 25)) +
  scale_y_continuous(limits = c(0, 45))

p3 <- data.frame(x = rpois(100, 2)) %>%
  ggplot(aes(x = x)) +
  geom_bar() +
  theme_bw() +
  theme(text = element_text(size = 25)) +
  labs(title = "Poisson(2)", x = "", y = "") +
  scale_x_continuous(limits = c(-1, 25)) +
  scale_y_continuous(limits = c(0, 45))

p4 <- data.frame(x = rpois(100, 5)) %>%
  ggplot(aes(x = x)) +
  geom_bar() +
  theme_bw() +
  theme(text = element_text(size = 25)) +
  labs(title = "Poisson(5)", x = "", y = "") +
  scale_x_continuous(limits = c(-1, 25)) +
  scale_y_continuous(limits = c(0, 45))

grid.arrange(p1, p2, p3, p4, ncol=2)
```

---

### Excess zeros

.pull-left[
.large[
.question[
*Why might there be excess 0s in the data, and why is that a problem for modeling the number of drinks consumed?*
]
]
]

.pull-right[
```{r echo=F, message=F, warning=F, fig.align='center', fig.width=6, fig.height=4}
wdrinks %>%
  ggplot(aes(x = drinks)) +
  geom_bar() +
  theme_bw() +
  theme(text = element_text(size = 25))
```
]

---

### Modeling

---

### Zero-inflated Poisson (ZIP) model

.large[
$$P(Y_i = y) = \begin{cases} e^{-\lambda_i}(1 - \alpha_i) + \alpha_i & y = 0 \\ \dfrac{e^{-\lambda_i} \lambda_i^y}{y!}(1 - \alpha_i) & y > 0 \end{cases}$$

where

$\log \left( \dfrac{\alpha_i}{1 - \alpha_i} \right) = \gamma_0 + \gamma_1 FirstYear_i + \gamma_2 OffCampus_i + \gamma_3 Male_i$

$\log(\lambda_i) = \beta_0 + \beta_1 FirstYear_i + \beta_2 OffCampus_i + \beta_3 Male_i$

]

---

### In R

```{r, output.lines = 10:22, message=F, warning=F}
library(pscl)
m1 <- zeroinfl(drinks ~ FirstYear + OffCampus + sex | 
                 FirstYear + OffCampus + sex, 
               data = wdrinks)
summary(m1)
```

---

### Interpretation

.large[
```{r, echo=F,output.lines = 17:22, message=F, warning=F}
summary(m1)
```

.question[
How would I interpret the estimated coefficient 0.892 in the logistic regression component of the model?
]
]

---

### Interpretation

.large[
```{r, echo=F,output.lines = 10:16, message=F, warning=F}
summary(m1)
```

.question[
How would I interpret the estimated coefficient 0.372 in the Poisson regression component of the model?
]
]