---
title: Exploring the Titanic data
output:
  xaringan::moon_reader:
    css: "lab-slides.css"
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
editor_options: 
  chunk_output_type: console
---

### What we've covered so far...

.large[
* Interpretation and model fitting (MLE, Fisher scoring, gradient ascent)
* Visualizations and diagnostics (empirical logit plots, quantile residual plots, VIFs, Cook's distance)
* Hypothesis testing (Wald tests)
]

---

### Data

.large[
Data on the RMS *Titanic* disaster. We have data on 891 passengers on the ship, with the following variables:

* `Passenger`: A unique ID number for each passenger.
* `Survived`: An indicator for whether the passenger survived (1) or perished (0) during the disaster.
* `Pclass`: Indicator for the class of the ticket held by this passengers; 1 = 1st class, 2 = 2nd class, 3 = 3rd class.
* `Sex`: Binary Indicator for the biological sex of the passenger.
* `Age`: Age of the passenger in years; Age is fractional if the passenger was less than 1 year old.
* `Fare`: How much the ticket cost in US dollars.
* `+` others
]

---

### Research question

.large[
*Is there a relationship between passenger age and their probability of survival, after accounting for sex, passenger class, and the cost of their ticket?*
]

.large[
.question[
What steps should I take to investigate this question with logistic regression?
]
]

---

### Class activity, Part I (EDA)

.large[
[https://sta712-f22.github.io/class_activities/ca_lecture_12.html](https://sta712-f22.github.io/class_activities/ca_lecture_12.html)

]

---

### Class activity

```{r include=F}
library(tidyverse)
titanic <- read.csv("https://sta214-f22.github.io/labs/Titanic.csv")

logodds_plot <- function(data, num_bins, bin_method,
                         x_name, y_name, grouping = NULL, 
                         reg_formula = y ~ x){
  
  if(is.null(grouping)){
    dat <- data.frame(x = data %>% pull(x_name), 
                      y = data %>% pull(y_name),
                      group = 1)
  } else {
    dat <- data.frame(x = data %>% pull(x_name), 
                      y = data %>% pull(y_name),
                      group = as.factor(data %>% pull(grouping)))
  }
  
  if(bin_method == "equal_size"){
    logodds_table <- dat %>%
      drop_na() %>%
      arrange(group, x) %>%
      group_by(group) %>%
      mutate(obs = y,
             bin = rep(1:num_bins,
                       each=ceiling(n()/num_bins))[1:n()]) %>%
      group_by(bin, group) %>%
      summarize(mean_x = mean(x),
                prop = mean(c(obs, 0.5)),
                num_obs = n()) %>%
      ungroup() %>%
      mutate(logodds = log(prop/(1 - prop)))
  } else {
    logodds_table <- dat %>%
      drop_na() %>%
      group_by(group) %>%
      mutate(obs = y,
             bin = cut(x, 
                       breaks = num_bins,
                       labels = F)) %>%
      group_by(bin, group) %>%
      summarize(mean_x = mean(x),
                prop = mean(c(obs, 0.5)),
                num_obs = n()) %>%
      ungroup() %>%
      mutate(logodds = log(prop/(1 - prop)))
  }
  
  if(is.null(grouping)){
    logodds_table %>%
      ggplot(aes(x = mean_x,
                 y = logodds)) +
      geom_point(size=2) +
      geom_smooth(se=F, method="lm", formula = reg_formula) +
      theme_bw() +
      labs(x = x_name,
           y = "Empirical log odds") +
      theme(text = element_text(size=15))
  } else {
    logodds_table %>%
      ggplot(aes(x = mean_x,
                 y = logodds,
                 color = group,
                 shape = group)) +
      geom_point(size=2) +
      geom_smooth(se=F, method="lm", formula = reg_formula) +
      theme_bw() +
      labs(x = x_name,
           y = "Empirical log odds",
           color = grouping,
           shape = grouping) +
      theme(text = element_text(size=15))
  }
  
}
```

```{r, echo=F, message=F, warning=F, fig.align='center', fig.width=10, fig.height=7}
p1 <- logodds_plot(titanic, 30, "equal_size", "Age", "Survived", 
             grouping = "Sex",
             reg_formula = y ~ x)

p2 <- logodds_plot(titanic, 30, "equal_size", "Fare", "Survived",
             reg_formula = y ~ x)

p3 <- logodds_plot(titanic, 30, "equal_size", "Fare", "Survived",
             reg_formula = y ~ log(x))

p4 <- logodds_plot(titanic, 30, "equal_size", "Fare", "Survived",
                   grouping = "Sex",
             reg_formula = y ~ log(x))

gridExtra::grid.arrange(p1, p2, p3, p4, ncol=2)
```

---

### Class activity

.large[
.question[
Based on your EDA, what model would you fit to address the research question?
]
]

---

### Class activity, Part II (Diagnostics)

.large[
[https://sta712-f22.github.io/class_activities/ca_lecture_12.html](https://sta712-f22.github.io/class_activities/ca_lecture_12.html)

]

---

### Class activity, Part III (Hypothesis testing)

.large[
[https://sta712-f22.github.io/class_activities/ca_lecture_12.html](https://sta712-f22.github.io/class_activities/ca_lecture_12.html)

]
