---
title: Overdispersion
output:
  xaringan::moon_reader:
    css: "lab-slides.css"
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
editor_options: 
  chunk_output_type: console
---

```{r include=F}
library(tidyverse)
library(knitr)
hook_output <- knit_hooks$get("output")
knit_hooks$set(output = function(x, options) {
   lines <- options$output.lines
   if (is.null(lines)) {
     return(hook_output(x, options))  # pass to default hook
   }
   x <- unlist(strsplit(x, "\n"))
   more <- "..."
   if (length(lines)==1) {        # first n lines
     if (length(x) > lines) {
       # truncate the output, but add ....
       x <- c(head(x, lines), more)
     }
   } else {
     x <- c(more, x[lines], more)
   }
   # paste these lines together
   x <- paste(c(x, ""), collapse = "\n")
   hook_output(x, options)
 })

crimes <- read_csv("~/Documents/Teaching/sta712-f22.github.io/slides/c_data.csv")

m2 <- glm(nv ~ region, offset = log(enroll1000),
          data = crimes, family = poisson)
summary(m2)


```


### Recap: Overdispersion

.large[
**Overdispersion** occurs when the response $Y$ has higher variance than we would expect from the specified EDM
]

---

### Estimating $\phi$

---

### Using $\widehat{\phi}$

.large[
```{r}
pearson_resids <- residuals(m2, type="pearson")
sum(pearson_resids^2)/df.residual(m2)
```

```{r echo=F, output.lines = 10:16}
summary(m2)
```
]

---

### Quasi-Poisson regression

.large[
A model for overdispersed Poisson-like counts, using an estimated dispersion parameter $\widehat{\phi}$, is called a *quasi-Poisson* model.

```{r, output.lines = 11:16}
m3 <- glm(nv ~ region, offset = log(enroll1000),
          data = crimes, family = quasipoisson)
summary(m3)
```
]

---

### Poisson vs. quasi-Poisson

.large[
**Poisson:**

```{r echo=F, output.lines=10:14}
summary(m2)
```

**Quasi-Poisson:**
```{r echo=F, output.lines=11:14}
summary(m3)
```
]

---

### Quasi-likelihood models
