---
title: Negative binomial regression
output:
  xaringan::moon_reader:
    css: "lab-slides.css"
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
editor_options: 
  chunk_output_type: console
---

```{r include=F}
library(tidyverse)
library(knitr)
library(gridExtra)
library(statmod)
hook_output <- knit_hooks$get("output")
knit_hooks$set(output = function(x, options) {
   lines <- options$output.lines
   if (is.null(lines)) {
     return(hook_output(x, options))  # pass to default hook
   }
   x <- unlist(strsplit(x, "\n"))
   more <- "..."
   if (length(lines)==1) {        # first n lines
     if (length(x) > lines) {
       # truncate the output, but add ....
       x <- c(head(x, lines), more)
     }
   } else {
     x <- c(more, x[lines], more)
   }
   # paste these lines together
   x <- paste(c(x, ""), collapse = "\n")
   hook_output(x, options)
 })

heart_data <- read.csv("~/Documents/Teaching/sta214-f22.github.io/class_activities/framingham.csv")

heart_data <- heart_data %>%
  drop_na(male, age, education, diabetes, cigsPerDay, BMI) %>%
  mutate(education = as.factor(education))

smokers <- heart_data %>%
  filter(currentSmoker == 1)

m1 <- glm(cigsPerDay ~ male + age + education + diabetes + BMI,
          data = smokers, family = poisson)

```

### Recap: negative binomial regression

.large[
$$Y_i \sim NB(r, \ p_i)$$

$$\log(\mu_i) = \beta^T X_i$$

* $\mu_i = \dfrac{p_i r}{1 - p_i}$
* Note that $r$ is the same for all $i$
* Note that just like in Poisson regression, we model the average count
  * Interpretation of $\beta$s is the same as in Poisson regression
]

---

### In R

.large[
```{r message=F}
library(MASS)
m2 <- glm.nb(cigsPerDay ~ male + age + education + 
               diabetes + BMI, data = smokers)
```

```{r echo=F, output.lines = c(11:17, 22, 23)}
summary(m2)
```

$\widehat{r} = 3.3$
]

---

### Class activity

.large[
[https://sta712-f22.github.io/class_activities/ca_lecture_27.html](https://sta712-f22.github.io/class_activities/ca_lecture_27.html)
]

---

### Class activity

```{r, echo=F, message=F, warning=F, fig.width=10, fig.height=7.5,fig.align='center'}
r <- 1
x <- rnorm(1000, mean=0, sd=1.2)
y1 <- rpois(1000, lambda = exp(x))
y2 <- rnbinom(1000, size=r, mu=exp(x))

sim_m1 <- glm(y1 ~ x, family = poisson)
sim_m2 <- glm.nb(y1 ~ x)

p1 <- data.frame(x = x, resids = qresid(sim_m1)) %>%
  ggplot(aes(x = x, y = resids)) +
  geom_point() +
  geom_smooth() +
  labs(x = "X", y = "Quantile residuals",
       title = "Poisson regression on Poisson data") +
  theme_bw()

p2 <- data.frame(x = x, resids = qresid(sim_m2)) %>%
  ggplot(aes(x = x, y = resids)) +
  geom_point() +
  geom_smooth() +
  labs(x = "X", y = "Quantile residuals",
       title = "Negative binomial regression on Poisson data") +
  theme_bw()

sim_m1 <- glm(y2 ~ x, family = poisson)
sim_m2 <- glm.nb(y2 ~ x)

p3 <- data.frame(x = x, resids = qresid(sim_m1)) %>%
  ggplot(aes(x = x, y = resids)) +
  geom_point() +
  geom_smooth() +
  labs(x = "X", y = "Quantile residuals",
       title = "Poisson regression on negative binomial data") +
  theme_bw()

p4 <- data.frame(x = x, resids = qresid(sim_m2)) %>%
  ggplot(aes(x = x, y = resids)) +
  geom_point() +
  geom_smooth() +
  labs(x = "X", y = "Quantile residuals",
       title = "Negative binomial regression on negative binomial data") +
  theme_bw()

grid.arrange(p1, p2, p3, p4, ncol=2)
```

---

### Poisson vs. negative binomial fits

```{r, echo=F, message=F, fig.align='center', fig.width=10, fig.height=4}
p1 <- smokers %>%
  mutate(qres = qresid(m1)) %>%
  ggplot(aes(x = age, y = qres)) +
  geom_point() +
  geom_smooth() +
  theme_bw() +
  labs(x = "Age", y = "Quantile residuals",
       title = "Poisson fit") +
  theme(text = element_text(size = 20))

p2 <- smokers %>%
  mutate(qres = qresid(m2)) %>%
  ggplot(aes(x = age, y = qres)) +
  geom_point() +
  geom_smooth() +
  theme_bw() +
  labs(x = "Age", y = "Quantile residuals",
       title = "Negative binomial fit") +
  theme(text = element_text(size = 20))

grid.arrange(p1, p2, ncol=2)
```

---

### Inference with negative binomial models

.large[
```{r echo=F, output.lines = c(11:19)}
summary(m2)
```

.question[
How would I test whether there is a relationship between age and the number of cigarettes smoked, after accounting for other variables?
]
]

---

### Inference with negative binomial models

.large[
```{r echo=F, output.lines = c(11:19)}
summary(m2)
```

.question[
How would I test whether there is a relationship between education and the number of cigarettes smoked, after accounting for other variables?
]
]

---

### Likelihood ratio test

.large[
```{r}
m2 <- glm.nb(cigsPerDay ~ male + age + education + 
               diabetes + BMI, data = smokers)
m3 <- glm.nb(cigsPerDay ~ male + age +
               diabetes + BMI, data = smokers)
m2$twologlik - m3$twologlik
pchisq(1.423, df=3, lower.tail=F)
```
]