---
title: Intro to mixed effects models
output:
  xaringan::moon_reader:
    css: "lab-slides.css"
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
editor_options: 
  chunk_output_type: console
---

### Motivating example: performance anxiety

.large[
We have data from a 2010 study on performance anxiety in 37 undergraduate music majors. For each musician, data was collected on anxiety levels before different performances (between 2 and 15 performances were measured for each musician), with variables including:
* `id`: a unique identifier for the musician
* `na`: negative affect score (a measure of anxiety)
* `perform_type`: whether the musican was performing in a large ensemble, small ensemble, or solo

.question[
How can we model the relationship between performance type and anxiety?
]

]

---

### A linear model for anxiety

.large[
$$Anxiety_i = \beta_0 + \beta_1 \ SmallEnsemble_i + \beta_2 \ Solo_i + \varepsilon_i$$

$\varepsilon_i \overset{iid}{\sim} N(0, \sigma_\varepsilon^2)$

.question[
What assumptions does this linear model make? Are all the assumptions reasonable?
]
]

---

### Exploratory data analysis

```{r, include=F}
library(tidyverse)
library(nnet)
library(gridExtra)
library(knitr)

hook_output <- knit_hooks$get("output")
knit_hooks$set(output = function(x, options) {
   lines <- options$output.lines
   if (is.null(lines)) {
     return(hook_output(x, options))  # pass to default hook
   }
   x <- unlist(strsplit(x, "\n"))
   more <- "..."
   if (length(lines)==1) {        # first n lines
     if (length(x) > lines) {
       # truncate the output, but add ....
       x <- c(head(x, lines), more)
     }
   } else {
     x <- c(more, x[lines], more)
   }
   # paste these lines together
   x <- paste(c(x, ""), collapse = "\n")
   hook_output(x, options)
 })

music <- read.csv("~/Documents/Teaching/sta214-f22.github.io/slides/musicdata.csv")
```

```{r, echo=F, message=F, warning=F, fig.align='center', fig.width=12, fig.height=5}
music %>% 
  ggplot(aes(x = factor(id), y = na)) + 
  geom_boxplot() + 
  facet_wrap(~perform_type) +
  theme_bw() +
  labs(x = "Musician", y = "Anxiety") +
  theme(text = element_text(size = 20))
```

.large[
.question[
Does it look like anxiety is correlated within musicians?
]
]

---

### Changing the model

.large[
$$Anxiety_i = \beta_0 + \beta_1 \ SmallEnsemble_i + \beta_2 \ Solo_i + \varepsilon_i$$

$\varepsilon_i \overset{iid}{\sim} N(0, \sigma_\varepsilon^2)$

.question[
How can we change the model to account for correlation within musicians?
]
]

---

### A mixed effects model

.large[
$$Anxiety_{ij} = \beta_0 + u_i + \beta_1 \ SmallEnsemble_{ij} + \beta_2 \ Solo_{ij} + \varepsilon_{ij}$$

$u_i \overset{iid}{\sim} N(0, \sigma_u^2) \hspace{1cm} \varepsilon_{ij} \overset{iid}{\sim} N(0, \sigma_\varepsilon^2)$
]

---

### Fitting the model in R

.large[
```{r, output.lines = c(11:15, 17:21), message=F, warning=F}
library(lme4)
m1 <- lmer(na ~ perform_type + (1|id), 
           data = music)
summary(m1)
```
]

---

### Assumptions

.large[
$$Anxiety_{ij} = \beta_0 + u_i + \beta_1 \ SmallEnsemble_{ij} + \beta_2 \ Solo_{ij} + \varepsilon_{ij}$$

$u_i \overset{iid}{\sim} N(0, \sigma_u^2) \hspace{1cm} \varepsilon_{ij} \overset{iid}{\sim} N(0, \sigma_\varepsilon^2)$
]

.large[
.question[
What assumptions does this mixed effects model make?
]
]
