---
title: ZIP models
output:
  xaringan::moon_reader:
    css: "lab-slides.css"
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
editor_options: 
  chunk_output_type: console
---

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=7, fig.height=5}
library(knitr)
library(pscl)
library(tidyverse)
library(statmod)
library(gridExtra)

hook_output <- knit_hooks$get("output")
knit_hooks$set(output = function(x, options) {
   lines <- options$output.lines
   if (is.null(lines)) {
     return(hook_output(x, options))  # pass to default hook
   }
   x <- unlist(strsplit(x, "\n"))
   more <- "..."
   if (length(lines)==1) {        # first n lines
     if (length(x) > lines) {
       # truncate the output, but add ....
       x <- c(head(x, lines), more)
     }
   } else {
     x <- c(more, x[lines], more)
   }
   # paste these lines together
   x <- paste(c(x, ""), collapse = "\n")
   hook_output(x, options)
 })

heart_data <- read.csv("~/Documents/Teaching/sta214-f22.github.io/class_activities/framingham.csv")

heart_data <- heart_data %>%
  drop_na(male, age, education, diabetes, cigsPerDay) %>%
  mutate(education = as.factor(education))
```

### Recap: Assessing the shape assumption

```{r, echo=F, message=F, warning=F, fig.align='center', fig.width=10, fig.height=7}

glm.pospois <- function(poisson_glm){
  x_mat <- model.matrix(poisson_glm)
  y <- poisson_glm$y
  
  y_pos <- y[y > 0]
  x_mat_pos <- x_mat[y > 0,]
  beta0 <- poisson_glm$coefficients
  
  U <- function(beta){
    l <- exp(c(x_mat_pos %*% beta))
    c(t(x_mat_pos) %*% (y_pos - l/(1 - exp(-l))))
  }
  
  I <- function(beta){
    l <- exp(c(x_mat_pos %*% beta))
    D <- diag(l * (1 - exp(-l)*(1 + l))/((1 - exp(-l))^2))
    t(x_mat_pos) %*% D %*% x_mat_pos
  }
  
  beta_old <- beta0
  for(i in 1:10){
    beta_new <- beta_old + solve(I(beta_old)) %*% U(beta_old)
    beta_old <- beta_new
  }
  
  output <- list("coefficients" = beta_new,
                 "y" = y_pos,
                 "x" = x_mat_pos)
  return(output)
  
}


## functions for quantile residual plots

## randomized quantile residuals for positive poisson data
## m = fitted positive poisson regression model
qresid_positive_poisson <- function(m){
  y <- m$y
  lambdas <- exp(m$x %*% m$coefficients)
  
  resids <- c()
  for(i in 1:length(y)){
    cdf_b <- (ppois(y[i], lambdas[i]) - ppois(0, lambdas[i]))/(1 - ppois(0, lambdas[i]))
    cdf_a <- (ppois((y[i]-1), lambdas[i]) - ppois(0, lambdas[i]))/(1 - ppois(0, lambdas[i]))
    resids[i] <- qnorm(runif(1, cdf_a, cdf_b))
  }
  return(resids)
}


## randomized quantile residuals for a ZIP model
## zip_m = fitted ZIP model from pscl
qresid_zeroinfl <- function(zip_m){
  y <- zip_m$y
  pred_probs <- predict(zip_m, type="prob")
  resids <- c()
  for(i in 1:length(y)){
    cdf_b <- sum(pred_probs[i,1:(y[i]+1)])
    if(y[i] == 0){
      cdf_a <- 0
    } else {
      cdf_a <- sum(pred_probs[i,1:y[i]])
    }
    
    resids[i] <- qnorm(runif(1, cdf_a, cdf_b))
  }
  return(resids)
}


# Now let's see how to do diagnostics for ZIP models

## Simulate some data where both the Poisson and logistic assumptions
## are met

x <- rnorm(1000)
alpha <- exp(x)/(1 + exp(x))
lambda <- exp(1 + x)
z <- rbinom(1000, 1, prob=alpha)
y <- 0*z + rpois(1000, lambda)*(1 - z)

## Fit a ZIP model and make a quantile residual plot
## Estimated coefficients are close to the true parameters
## The plot looks good

m <- zeroinfl(y ~ x | x)
# summary(m)

p1 <- data.frame(x = x, resids = qresid_zeroinfl(m)) %>%
  ggplot(aes(x = x, y = resids)) +
  geom_point() +
  geom_smooth() +
  theme_bw() +
  labs(title = "All assumptions satisfied")




## Now simulate data where the Poisson assumption is violated

x <- rnorm(1000)
alpha <- exp(x)/(1 + exp(x))
lambda <- exp(1 + x + 0.5*x^2)
z <- rbinom(1000, 1, prob=alpha)
y <- 0*z + rpois(1000, lambda)*(1 - z)

## Fit a ZIP model and make a quantile residual plot
## Estimated coefficients are wrong
## The plot looks bad

m <- zeroinfl(y ~ x | x)
# summary(m)

p2 <- data.frame(x = x, resids = qresid_zeroinfl(m)) %>%
  ggplot(aes(x = x, y = resids)) +
  geom_point() +
  geom_smooth() +
  theme_bw() +
  labs(title = "Poisson shape assumption violated")


## Finally, simulate data where the logistic assumption is violated
## Quantile residual plot doesn't look right!

x <- rnorm(1000)
alpha <- exp(-2*x^2)/(1 + exp(-2*x^2))
lambda <- exp(1 + x)
z <- rbinom(1000, 1, prob=alpha)
y <- 0*z + rpois(1000, lambda)*(1 - z)

m <- zeroinfl(y ~ x | x)
# summary(m)

p3 <- data.frame(x = x, resids = qresid_zeroinfl(m)) %>%
  ggplot(aes(x = x, y = resids)) +
  geom_point() +
  geom_smooth() +
  theme_bw() +
  labs(title = "Logistic shape assumption violated")

grid.arrange(p1, p2, p3, ncol=2)
```


---

### Logistic component vs. Poisson component

---

### Class activity

.large[
[https://sta712-f22.github.io/class_activities/ca_lecture_33.html](https://sta712-f22.github.io/class_activities/ca_lecture_33.html)
]

---

### Class activity

.large[
$$\log \left( \dfrac{\alpha_i}{1 - \alpha_i} \right) = \gamma_0 + \gamma_1 EducationSome_i + \gamma_2 EducationCollege_i + \\ \gamma_3 EducationAdv_i + \gamma_4 Diabetes_i + \gamma_5 Age_i$$

$$\log(\lambda_i) = \beta_0 + \beta_1 EducationSome_i + \beta_2 EducationCollege_i + \\ \beta_3 EducationAdv_i + \beta_4 Diabetes_i + \beta_5 Age_i$$

.question[
Research question: for smokers, does the number of cigarettes smoked per day depend on age?

What are the null and alternative hypotheses?
]
]

---

### Class activity

.large[
$$\log \left( \dfrac{\alpha_i}{1 - \alpha_i} \right) = \gamma_0 + \gamma_1 EducationSome_i + \gamma_2 EducationCollege_i + \\ \gamma_3 EducationAdv_i + \gamma_4 Diabetes_i + \gamma_5 Age_i$$

$$\log(\lambda_i) = \beta_0 + \beta_1 EducationSome_i + \beta_2 EducationCollege_i + \\ \beta_3 EducationAdv_i + \beta_4 Diabetes_i + \beta_5 Age_i$$

.question[
Research question: is there a relationship between age and whether someone is a smoker?

What are the null and alternative hypotheses?
]
]

---

### Wald tests

.large[
.question[
Research question: is there a relationship between age and whether someone is a smoker?
]
]

```{r, output.lines = 19:26}
m1 <- zeroinfl(cigsPerDay ~ education + diabetes + 
                 age | education + diabetes + age, 
               data = heart_data)
summary(m1)
```

---

### Class activity

.large[
$$\log \left( \dfrac{\alpha_i}{1 - \alpha_i} \right) = \gamma_0 + \gamma_1 EducationSome_i + \gamma_2 EducationCollege_i + \\ \gamma_3 EducationAdv_i + \gamma_4 Diabetes_i + \gamma_5 Age_i$$

$$\log(\lambda_i) = \beta_0 + \beta_1 EducationSome_i + \beta_2 EducationCollege_i + \\ \beta_3 EducationAdv_i + \beta_4 Diabetes_i + \beta_5 Age_i$$

.question[
Research question: Is there a relationship between education level and the number of cigarettes smoked?

What are the null and alternative hypotheses?
]
]

---

### Likelihood ratio test

.large[
```{r}
m1 <- zeroinfl(cigsPerDay ~ education + diabetes + 
                 age | education + diabetes + age, 
               data = heart_data)
m2 <- zeroinfl(cigsPerDay ~ education + diabetes 
               | education + diabetes, 
               data = heart_data)

2*(m1$loglik - m2$loglik)

pchisq(242.281, df=6, lower.tail=F)
```
]
