---
title: Intro to mixed effects models
output:
  xaringan::moon_reader:
    css: "lab-slides.css"
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
editor_options: 
  chunk_output_type: console
---

```{r, include=F}
library(tidyverse)
library(nnet)
library(gridExtra)
library(knitr)

hook_output <- knit_hooks$get("output")
knit_hooks$set(output = function(x, options) {
   lines <- options$output.lines
   if (is.null(lines)) {
     return(hook_output(x, options))  # pass to default hook
   }
   x <- unlist(strsplit(x, "\n"))
   more <- "..."
   if (length(lines)==1) {        # first n lines
     if (length(x) > lines) {
       # truncate the output, but add ....
       x <- c(head(x, lines), more)
     }
   } else {
     x <- c(more, x[lines], more)
   }
   # paste these lines together
   x <- paste(c(x, ""), collapse = "\n")
   hook_output(x, options)
 })

music <- read.csv("~/Documents/Teaching/sta214-f22.github.io/slides/musicdata.csv")
```

### Recap: data and motivation

.large[
We have data from a 2010 study on performance anxiety in 37 undergraduate music majors. For each musician, data was collected on anxiety levels before different performances (between 2 and 15 performances were measured for each musician), with variables including:
* `id`: a unique identifier for the musician
* `na`: negative affect score (a measure of anxiety)
* `perform_type`: whether the musican was performing in a large ensemble, small ensemble, or solo

.question[
How can we model the relationship between performance type and anxiety?
]

]

---

### Recap: a mixed effects model

.large[
$$Anxiety_{ij} = \beta_0 + u_i + \beta_1 \ SmallEnsemble_{ij} + \beta_2 \ Solo_{ij} + \varepsilon_{ij}$$

$u_i \overset{iid}{\sim} N(0, \sigma_u^2) \hspace{1cm} \varepsilon_{ij} \overset{iid}{\sim} N(0, \sigma_\varepsilon^2)$
]

---

### Fitting the model in R

.large[
```{r, output.lines = c(11:15, 17:21), message=F, warning=F}
library(lme4)
m1 <- lmer(na ~ perform_type + (1|id), 
           data = music)
summary(m1)
```
]

---

### Interpretation

.large[
```{r, echo=F, output.lines = c(17:21), message=F, warning=F}
summary(m1)
```

.question[
How would we interpret the estimated fixed effects?
]
]

---

### Prediction

.large[
```{r, echo=F, output.lines = c(17:21), message=F, warning=F}
summary(m1)
```

.question[
What is the estimated anxiety for Musician 1 before a solo performance? 
]
]

---

### Prediction

.large[
.question[
What is the estimated anxiety for Musician 1 before a solo performance? 
]
]

```{r, output.lines = c(1:5), message=F, warning=F}
coef(m1)
```


---

### Prediction

.large[
```{r, echo=F, output.lines = c(17:21), message=F, warning=F}
summary(m1)
```

.question[
What is the estimated anxiety for a *new* musician (not in the data) before a solo performance? 
]
]

---

### Assumptions

.large[
$$Anxiety_{ij} = \beta_0 + u_i + \beta_1 \ SmallEnsemble_{ij} + \beta_2 \ Solo_{ij} + \varepsilon_{ij}$$

$u_i \overset{iid}{\sim} N(0, \sigma_u^2) \hspace{1cm} \varepsilon_{ij} \overset{iid}{\sim} N(0, \sigma_\varepsilon^2)$
]

.large[
.question[
What assumptions does this mixed effects model make?
]
]

---

### Assessing normality

.large[
$$Anxiety_{ij} = \beta_0 + u_i + \beta_1 \ SmallEnsemble_{ij} + \beta_2 \ Solo_{ij} + \varepsilon_{ij}$$

$u_i \overset{iid}{\sim} N(0, \sigma_u^2) \hspace{1cm} \varepsilon_{ij} \overset{iid}{\sim} N(0, \sigma_\varepsilon^2)$
]

.large[
.question[
How should we check the normality assumption?
]
]

---

### QQ plots

.pull-left[
```{r message=F, warning=F, echo=F, fig.align='center', fig.height=4, fig.width=6}

music %>%
  mutate(resids = residuals(m1)) %>%
  ggplot(aes(sample = resids)) +
  geom_qq(size=2) +
  geom_qq_line(color="blue", lwd=1.2) +
  labs(x = "Theoretical normal quantiles", 
       y = "Observed residual quantiles") +
  theme_bw() +
  theme(text = element_text(size = 15))
```
]

.pull-right[
```{r message=F, warning=F, echo=F, fig.align='center', fig.height=4, fig.width=6}

data.frame(re = coef(m1)$id[,1] - summary(m1)$coefficients[1,1]) %>%
  ggplot(aes(sample = re)) +
  geom_qq(size=2) +
  geom_qq_line(color="blue", lwd=1.2) +
  labs(x = "Theoretical normal quantiles", 
       y = "Observed random effect quantiles") +
  theme_bw() +
  theme(text = element_text(size = 15))
```
]

---

### Changing the model

.large[
$$Anxiety_{ij} = \beta_0 + u_i + \beta_1 \ SmallEnsemble_{ij} + \beta_2 \ Solo_{ij} + \varepsilon_{ij}$$

$u_i \overset{iid}{\sim} N(0, \sigma_u^2) \hspace{1cm} \varepsilon_{ij} \overset{iid}{\sim} N(0, \sigma_\varepsilon^2)$
]

.large[
.question[
How could we change the model to allow the effect of performance type to differ between musicians?
]
]

---

### Fitting the model

```{r, output.lines = c(11:23), message=F, warning=F}
m2 <- lmer(na ~ perform_type + (perform_type|id), 
           data = music)
summary(m2)
```

---

### Prediction

.large[
.question[
What is the estimated anxiety for Musician 1 before a solo performance? 
]
]

```{r, output.lines = c(1:5), message=F, warning=F}
coef(m2)
```

