---
title: "Class Activity, October 21"
output: 
  rmdformats::robobook:
    css: "homework.css"
    highlight: pygments
link-citations: yes
---

# Exploring Poisson regression

In this activity, we are interested in analyzing the number of articles published by biochemistry PhD students. The data contains the following variables:

* `art`: articles published in last three years of Ph.D.
* `fem`: gender (recorded as `Men` or `Women` in the data)
* `mar`: marital status (recorded as married or single)
* `kid5`: number of children under age six
* `phd`: prestige of Ph.D. program
* `ment`: articles published by their mentor in last three years

```{r, eval=F}
library(foreign)

articles <- read.dta("http://www.stata-press.com/data/lf2/couart2.dta")
```

## Questions

Since the number of articles published is a count variable, your friend proposes the following Poisson regression model:

$$Articles_i \sim Poisson(\lambda_i)$$

$$\log(\lambda_i) = \beta_0 + \beta_1 Female_i + \beta_2 Married_i + \beta_3 Kids_i + \\  \beta_4 Prestige_i + \beta_5 Mentor_i$$

### EDA

1. Do you think we need to include an offset in our model? If so, what would the offset be?

Now let's explore our data. Our Poisson regression model assumes that the log mean number of articles published has a linear relationship with the explanatory variables. We can investigate whether this assumption seems reasonable by creating an *empirical log means* plot, which is analogous to our empirical logits plot from logistic regression. The following function can be used to create an empirical log means plot:

```{r eval=F}
logmean_plot <- function(data, num_bins, bin_method,
                         x, y, grouping = NULL, reg_formula = y ~ x){
  
  if(is.null(grouping)){
    dat <- data.frame(x = data[,x], 
                      y = data[,y],
                      group = 1)
  } else {
    dat <- data.frame(x = data[,x], 
                      y = data[,y],
                      group = data[,grouping])
  }
  
  if(bin_method == "equal_size"){
    log_table <- dat %>%
      drop_na() %>%
      arrange(group, x) %>%
      group_by(group) %>%
      mutate(obs = y,
             bin = rep(1:num_bins,
                       each=ceiling(n()/num_bins))[1:n()]) %>%
      group_by(bin, group) %>%
      summarize(mean_x = mean(x),
                mean_y = mean(obs),
                num_obs = n()) %>%
      ungroup() %>%
      mutate(log_mean = log(mean_y))
  } else {
    log_table <- dat %>%
      drop_na() %>%
      group_by(group) %>%
      mutate(obs = y,
             bin = cut(x, 
                       breaks = num_bins,
                       labels = F)) %>%
      group_by(bin, group) %>%
      summarize(mean_x = mean(x),
                mean_y = mean(obs),
                num_obs = n()) %>%
      ungroup() %>%
      mutate(log_mean = log(mean_y))
  }
  
  if(is.null(grouping)){
    log_table %>%
      ggplot(aes(x = mean_x,
                 y = log_mean)) +
      geom_point(size=2.5) +
      geom_smooth(se=F, method="lm", formula = reg_formula) +
      theme_bw() +
      labs(x = x,
           y = "Empirical log mean count") +
      theme(text = element_text(size=25))
  } else {
    log_table %>%
      ggplot(aes(x = mean_x,
                 y = log_mean,
                 color = group,
                 shape = group)) +
      geom_point(size=2.5) +
      geom_smooth(se=F, method="lm", formula = reg_formula) +
      theme_bw() +
      labs(x = x,
           y = "Empirical log mean count",
           color = grouping,
           shape = grouping) +
      theme(text = element_text(size=25))
  }
  
}
```

2. Use the code below to investigate the relationship between the log mean number of articles, and the prestige of the PhD program, and experiment with a few different numbers of bins. Do you think a transformation will be needed for Prestige?

```{r, eval=F}
logmean_plot(articles, 20, "equal_size", "phd", "art")
```

3. Explore the relationship between the log mean number of articles, and the number of articles the student's mentor has published. Do you think a transformation is needed for Mentor?

4. Use the `grouping` argument in the `logmean_plot` function to assess whether we should include interaction terms in the model.

5. Based on your initial data exploration, would you modify the model your friend suggested in any way? If so, write down your chosen model.

### Model fitting and interpretation

6. Using the `glm` function in R (with `family = poisson`), fit your model from Question 5. 

7. Interpret the fitted coefficient on Prestige.

8. Using your fitted model, what is the estimated number of articles published in the last three years for a married, male student with 0 children, attending a PhD program with $Prestige = 3$, whose advisor has published 10 articles in the last 3 years?

### Diagnostics

9. 