---
title: "Class Activity, October 31"
output: 
  rmdformats::robobook:
    css: "homework.css"
    highlight: pygments
link-citations: yes
---

# Negative binomial vs. Poisson regression

The negative binomial distribution allows a more flexible mean-variance relationship than the Poisson distribution. For the Poisson distribution, $V(\mu) = \mu$, whereas for the negative binomial $V(\mu) = \mu + \dfrac{\mu^2}{r}$. For large values of $r$, the negative binomial is similar to the Poisson, whereas for small values of $r$ the negative binomial allows much more variability in the response.

In this class activity, we will simulate data from the Poisson and the negative binomial, and we will compare Poisson regression with negative binomial regression.

### Simulating data

Run the following code in R to simulate $X$ from a normal distribution, and two different response variables; $Y_{(1)}$ is Poisson, and $Y_{(2)}$ is negative binomial with $r = 1$.

```{r, eval=F}
r <- 1
x <- rnorm(1000, mean=0, sd=1.2)
y1 <- rpois(1000, lambda = exp(x))
y2 <- rnbinom(1000, size=r, mu=exp(x))
```

### Visualizing the results

Let's compare the plots of $Y_{(1)}$ vs. $X$ and $Y_{(2)}$ vs $X$. Run the following code:

```{r, eval=F}
plot(x, y1, main = "Poisson response")
plot(x, y2, main = "Negative binomial response")
```

**Question 1:** What do you notice about the difference in the variability of $Y$ for the Poisson vs. negative binomial data?

### Fitting models on the Poisson data

Now let's fit both a Poisson regression model and a negative binomial regression model *on the Poisson data*, and compare the quantile residual plots.


```{r, eval=F}
library(MASS)
library(statmod)
library(tidyverse)

m1 <- glm(y1 ~ x, family = poisson)
m2 <- glm.nb(y1 ~ x)

summary(m1)
summary(m2)

data.frame(x = x, resids = qresid(m1)) %>%
  ggplot(aes(x = x, y = resids)) +
  geom_point() +
  geom_smooth() +
  labs(x = "X", y = "Quantile residuals",
       title = "Poisson regression on Poisson data") +
  theme_bw()

data.frame(x = x, resids = qresid(m2)) %>%
  ggplot(aes(x = x, y = resids)) +
  geom_point() +
  geom_smooth() +
  labs(x = "X", y = "Quantile residuals",
       title = "Negative binomial regression on Poisson data") +
  theme_bw()
```

**Question 2:** Are the estimated coefficients for the two models similar? What is the estimate $\widehat{r}$ for the negative binomial model?

**Question 3:** Do the quantile residual plots for both models look reasonable?


### Fitting models on the negative binomial data

Now let's fit both a Poisson regression model and a negative binomial regression model *on the negative binomial data*, and compare the quantile residual plots.


```{r, eval=F}
m1 <- glm(y2 ~ x, family = poisson)
m2 <- glm.nb(y2 ~ x)

summary(m1)
summary(m2)

data.frame(x = x, resids = qresid(m1)) %>%
  ggplot(aes(x = x, y = resids)) +
  geom_point() +
  geom_smooth() +
  labs(x = "X", y = "Quantile residuals",
       title = "Poisson regression on negative binomial data") +
  theme_bw()

data.frame(x = x, resids = qresid(m2)) %>%
  ggplot(aes(x = x, y = resids)) +
  geom_point() +
  geom_smooth() +
  labs(x = "X", y = "Quantile residuals",
       title = "Negative binomial regression on negative binomial data") +
  theme_bw()
```

**Question 4:** Are the estimated coefficients for the two models similar? What is the estimate $\widehat{r}$ for the negative binomial model?

**Question 5:** Do the quantile residual plots for both models look reasonable?

