---
title: "Class activity solutions, September 26"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

### Question 1

```{r message=F, warning=F}
# read in the data, and convert passenger class with a factor
# specify that the order is third, second, first
titanic <- read.csv("https://sta214-f22.github.io/labs/Titanic.csv")
titanic <- titanic %>%
  drop_na() %>%
  mutate(Pclass = factor(Pclass, levels = c(3, 2, 1))) 

# fit a logistic regression model
m1 <- glm(Survived ~ Sex*Age + Pclass, data = titanic, family = binomial)

# now we want to construct a confidence interval for beta_4 - beta_3
# first, create a vector to specify the linear combination of coefficients
# (note that by default, R puts interaction coefficients at the end)
a <- c(0, 0, 0, -1, 1, 0)

# calculate lower and upper bounds
t(a) %*% coef(m1) - qnorm(0.975) * sqrt(t(a) %*% vcov(m1) %*% a)
t(a) %*% coef(m1) + qnorm(0.975) * sqrt(t(a) %*% vcov(m1) %*% a)
```

We are 95% confident that the true difference in log odds of survival between first and second class passengers, holding sex and age fixed, is between 0.913 and 2.098.

Since log odds are hard to interpret, we might prefer to interpret our confidence interval on the odds scale instead. Since $e^{0.913} = 2.492$ and $e^{2.098} = 8.150$, we are 95% confident that the odds of survival for a first class passenger are between 2.492 and 8.150 times higher than the odds of survival for a second class passenger of the same age and sex.


### Question 2

Here we need to modify the code to construct a confidence interval for the log odds of survival for a female passenger, aged 20, in second class. Plugging in to our model, the log odds of survival would be
$$\beta_0 + \beta_1 (0) + \beta_2 (20) + \beta_3 (1) + \beta_4(0) + \beta_5(0)$$

so our linear combination vector is $a^T = (1, 0, 20, 1, 0, 0)$.

```{r message=F, warning=F}
a <- c(1, 0, 20, 1, 0, 0)

# calculate lower and upper bounds
t(a) %*% coef(m1) - qnorm(0.975) * sqrt(t(a) %*% vcov(m1) %*% a)
t(a) %*% coef(m1) + qnorm(0.975) * sqrt(t(a) %*% vcov(m1) %*% a)
```

We are 95% confident that the log odds of survival for a 20 year old female passenger in second class are between 0.982 and 1.931.

### Question 3

To calculate the a confidence interval for the *probability* of survival, we can simply convert the endpoints of our interval from Question 2 into probabilities.

$\dfrac{e^{0.982}}{1 + e^{0.982}} = 0.728$

$\dfrac{e^{1.931}}{1 + e^{1.931}} = 0.873$

We are 95% confident that the probability of survival for a 20 year old female passenger in second class is between 0.728 and 0.873.