---
title: "Class Activity, November 16"
output: 
  rmdformats::robobook:
    css: "homework.css"
    highlight: pygments
link-citations: yes
---

# Modeling earthquake damage

We have data from the 2015 Gorkha earthquake in Nepal. After the earthquake, a large scale survey was conducted to determine the amount of damage the earthquake caused for homes, businesses and other structures. Variables include:

* `Damage`: the amount of damage suffered by the building (none, moderate, severe)
* `age`: the age of the building (in years)
* `condition`: a de-identified variable recording the condition of the land surrounding the building (n, o, or t)

We fit a multinomial regression model to predict Damage:

$$Damage_i \sim Categorical(\pi_{i(None)}, \pi_{i(Moderate)}, \pi_{i(Severe)})$$
$$\log \left( \dfrac{\pi_{i(Moderate)}}{\pi_{i(None)}} \right) = \beta_{0(Moderate)} + \beta_{1(Moderate)} Age_i^{1/2} + \\ \hspace{5cm} \beta_{2(Moderate)} ConditionO_i + \beta_{3(Moderate)} ConditionT_i$$
$$\log \left( \dfrac{\pi_{i(Severe)}}{\pi_{i(None)}} \right) = \beta_{0(Severe)} + \beta_{1(Severe)} Age_i^{1/2} +  \\ \hspace{5cm} \beta_{2(Severe)} ConditionO_i + \beta_{3(Severe)} ConditionT_i$$


The output of the fitted model is shown below:

```{r include=F}
library(nnet)
library(tidyverse)
earthquake <- read.csv("https://sta214-f22.github.io/labs/EarthquakeData.csv")

set.seed(100)

earthquake <- earthquake %>%
  slice_sample(n=10000) %>%
  dplyr::select(Damage, count_families, 
         foundation_type, land_surface_condition, 
         age) %>%
  mutate(Damage = factor(Damage, levels = c("none", "moderate", "severe"))) %>%
  rename(condition = land_surface_condition)

m1 <- multinom(Damage ~ sqrt(age) + 
                 condition, 
               data = earthquake)
```

```{r, eval=F}
m1 <- multinom(Damage ~ sqrt(age) + 
                 condition, 
               data = earthquake)
```

```{r}
summary(m1)
```

## Questions

1. What is the predicted odds of moderate damage vs. no damage for a 25 year old building with surface condition O?

2. What is the predicted *probability* of moderate damage for a 25 year old building with surface condition O?